Bayesian inference of material culture phylogenies using continuous traits: A birth–death model for Late Neolithic/Early Bronze Age arrowheads from Northwestern Europe

### 2023-10-20 analysis/paper/paper.qmd
minor edits on submission to PCI Archaeology

In palaeontological phylogenetics there are three likelihood models that are the starting points for most analyses: the substitution model, the clock model, and the tree model [@warnockUnderstandingTripartiteApproach2020; @fig-tripartite-models]. Substitution models describe the process the researcher believes generated the character data. There are two main categories, each having several variants: discrete (e.g. Mk) and continuous (e.g. Brownian motion, Ornstein--Uhlenbeck, Lévy processes). Clock models describes the manner in which the researcher believes rates of evolution are distributed across the tree. There are two main groups of clock models: strict (with a constant rate among branches) and relaxed, with examples including uncorrelated relaxed clock (with a distinct rate along each branch drawn from a chosen probability distribution), and autocorrelated relaxed clock (with a distinct rate along each branch that is correlated with the rate along its parent branch). Tree models describe the extinction, speciation, and sampling events that potentially led to the observed trees. Examples of tree models include pure birth (assumes speciation is constant over time, that we have no extinction), birth-death (where are species or variant disappear through extinction), and fossilized birth-death trees (incorporates the fossil recovery process and provides an expression for the probability of observing a tree with samples recovered along internal branches). The tripartite model is an important framework that enables nearly endless combinations of substitution, clock, and tree models to be assembled into a complete model.

@@

```{r}
#| label: fig-efa-schematic
#| fig-cap: "Schematic visualisation of the workflow proposed and used in this study. First, the outlines of 2D artefact images are described in so called harmonics using Elliptic Fourier Analysis, then these harmonics are then transformed and reduced in dimensions using a Principal Components Analysis. The Principal Components can be interpreted as artefact shape variables and be used as continuous character traits for the inference of phylogenies."

@@

Landmark and outline analysis of artefact shape is widely used in archaeology. Landmark-based morphometrics has been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. For artefacts where landmarks are ambiguous or absent, outline approaches, such as Elliptic Fourier Analysis (EFA), are more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Bonhomme2014]. Coordinates along a curve are converted into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. Outline analyses has long been effectively employed in the analysis of stone artefacts [e.g. @Gero1984; @Iovictua2009; @Fox2015; @Lipo2016; @Hoggard2019; @Ivanovaite2020; @riede2019reconciling; @matzig2021] and pottery [@Wilczek2014; @wang2020]. These studies share a standard approach of digitizing artefact shapes, extracting outlines, computing an EFA on these outlines, and then reducing the dimensionality of the harmonics by inputting them into a principal component analysis (PCA). The principal component (PC) scores are then interpreted as artefact shape variables that are typically more informative than linear dimensions. In our case study we extend the use of these artefact shape variables by using them as continuous characters to input into phylogenetic models [@fig-efa-schematic].

@@

In RevBayes, we generated fours models to investigate the most appropriate combination of substitution, clock, and tree models. @tbl-model-summary summarises the model configurations and their marginal likelihoods. Full details of our Bayesian MCMC procedure are documented in our supplementary materials at http://doi.org/xxx/xxx. Taking the first model as the null model, the Bayes Factors indicate strong support for the model using a birth-death tree (where variants can appear and disappear, rather than only appear), relaxed Brownian motion (where the rate of morphological change is allowed to vary among branches of the phylogeny) and a relaxed clock (allowing for shifts in rates of morphological change across branches). @fig-treemap-densitree-map summarises the phylogenetic landscape, with the top left panel showing a single cluster of related trees, indicating that one phylogeny is definitively supported by the data [@jombartTreespaceStatisticalExploration2017]. The top right panel is a Densitree plot of trees randomly sampled from the converged posterior to provide intuition about the extent of incongruence among these trees [@bouckaertDensiTreeMakingSense2010]. It shows high congruence for the Bronze A2 and Bell Beaker samples, but substantial uncertainty for the Bronze A1 arrowheads, indicating competing topologies and possibly horizontal transmission and independent convergence on similar shapes.

>>>>

In palaeontological phylogenetics there are three likelihood models that are the starting points for most analyses (@fig-tripartite-models): the substitution model, the clock model, and the tree model [@warnock2020; @Ho_Duchene_2014]. Substitution models describe the processes we believe produced the character data. There are two main categories, each having several variants: discrete (e.g. Mk) and continuous (e.g. Brownian motion, Ornstein--Uhlenbeck, Lévy processes). Clock models describe our beliefs about how rates of evolution are distributed across a tree. There are two main groups of clock models: strict (with a constant rate among branches) and relaxed, with examples including uncorrelated relaxed clock (with a distinct rate along each branch drawn from a probability distribution), and autocorrelated relaxed clock (with a distinct rate along each branch that is correlated with the rate along its parent branch) [@Ho_Duchene_2014]. Tree models describe the extinction, diversification, and sampling events that may have produced the observed trees. Examples of tree models include pure birth (assumes diversification is constant over time, that we have no extinction), birth-death (where variants disappear through extinction), and fossilized birth-death trees (incorporates the fossil recovery process and gives the probability of observing a tree with samples recovered along its internal branches) [@warnock2020]. The tripartite model is an important framework that allows a complete model to be assembled from combinations of substitution, clock, and tree models [@warnock2020; @Ho_Duchene_2014].

@@

```{r}
#| label: fig-efa-schematic
#| fig-cap: "Schematic visualisation of the workflow proposed and used in this study. First, the outlines of 2D artefact images are described in so called harmonics using Elliptic Fourier Analysis, then these harmonics are then transformed and reduced in dimensions using a Principal Components Analysis. The Principal Components can be interpreted as artefact shape variables and be used as continuous character traits for the inference of phylogenies. Modified from original figures by Josef Wilczek and Jonathan R. Hendricks."

@@

Landmark and outline analysis of artefact shape is widely used in archaeology. Landmark-based morphometrics has been productively employed using archaeological artefacts that have morphological features that act as unambiguous reference points for measurement. These reference points have included tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone morphologies [@Haruda2019], or ceramics with distinct features [@Selden2019; @Topi2017]. For artefacts where landmarks are ambiguous or absent, approaches that use an object's outline shape, such as Elliptic Fourier Analysis (EFA), can be effective for assessing morphological variations [@Cardillo2010]. EFA decomposes the outline into a series of ellipses, and describes those ellipses with a series of trigonometric functions [@Bonhomme2014]. Coordinates located along the curves of an object's outline are converted into Fourier function coefficients, also known as harmonic coefficients or harmonics [@Kuhl1982]. The the quality and precision of the geometric representation of an object is determined by the number of harmonics we keep for further analysis. Outline analyses has long been effectively employed in the analysis of stone artefacts [e.g. @Gero1984; @Iovictua2009; @Fox2015; @Lipo2016; @Hoggard2019; @Ivanovaite2020; @riede2019reconciling; @matzig2021] and pottery [@Wilczek2014; @wang2020]. These studies share a standard approach of digitizing artefact shapes, extracting outlines, computing an EFA on these outlines, and then reducing the dimensionality of the harmonics by inputting them into a principal component analysis (PCA). The principal component (PC) scores are then interpreted as artefact shape variables that are typically more informative than linear dimensions. In our case study we extend the use of these artefact shape variables by using them as continuous characters to input into phylogenetic models [@fig-efa-schematic].

@@

In RevBayes, we generated fours models to investigate the most appropriate combination of substitution, clock, and tree models. @tbl-model-summary summarises the model configurations and their marginal likelihoods. Full details of our Bayesian MCMC procedure are documented in our supplementary materials at http://doi.org/10.17605/OSF.IO/RBQA5 Taking the first model as the null model, the Bayes Factors indicate strong support for the model using a birth-death tree (where variants can appear and disappear, rather than only appear), relaxed Brownian motion (where the rate of morphological change is allowed to vary among branches of the phylogeny) and a relaxed clock (allowing for shifts in rates of morphological change across branches). @fig-treemap-densitree-map summarises the phylogenetic landscape, with the top left panel showing a single cluster of related trees, indicating that one phylogeny is definitively supported by the data [@jombartTreespaceStatisticalExploration2017]. The top right panel is a Densitree plot of trees randomly sampled from the converged posterior to provide intuition about the extent of incongruence among these trees [@bouckaertDensiTreeMakingSense2010]. It shows high congruence for the Bronze A2 and Bell Beaker samples, but substantial uncertainty for the Bronze A1 arrowheads, indicating competing topologies and possibly horizontal transmission and independent convergence on similar shapes.
